The Hyperedge Event Model

### 2018-06-20 paper/ba-HEM/ba-sample.tex
addressed remianing comments

		\begin{abstract}
			We introduce the hyperedge event model (HEM)---a generative model for events that can be represented as directed edges with one sender and one or more receivers or one receiver and one or more senders. To define the model, we
			integrate a dynamic version of the exponential random graph model (ERGM) of edge structure with a survival model for event timing to jointly understand who interacts with whom, and when. The HEM offers three innovations with respect to the literature---first, it extends a growing class of dynamic network models to model hyperedges. Second, we introduce a novel receiver-selection distribution that is natural for generating sets of recipients. Third, we incorporate both a timing equation and an edge-formation equation into the model specification. We use the HEM to analyze emails sent among department managers in Montgomery County government in North Carolina. Our application demonstrates that the model is effective at predicting and explaining time-stamped network data involving edges with multiple receivers. We present an out-of-sample prediction experiment to illustrate how researchers can select between different specifications of the model. 
		\end{abstract}

@@

	In this section we define the HEM by describing a process according to which hyperedge events are generated in continuous time. Data generated under the HEM consists of $D$ unique edges. A single edge indexed by $d \in [D]$, where $[D]$ denotes a categorical set with $D$ levels $[D] = \{1,\ldots,D\}$, is represented by the three components: the sender $a_d \in [A]$, an indicator vector of receivers $\boldsymbol{r}_d = (u_{dr})_{r=1}^{A}$, and the timestamp $t_d \in (0, \infty)$. For simplicity, we assume that edges are ordered by time such that $t_d \leq t_{d+1}$. While the model can be applied for two types of hyperedges---edges with (1) one sender and one or more receivers, and (2) one or more senders and one receiver---here we only present the generative process for those involving one sender and one or more receivers (i.e., multicast). One notable feature of the HEM generative process is that we draw auxiliary variables that serve as candidate data. Data is generated from the HEM through a sampling process applied to the auxiliary variables. The auxiliary variables drawn for edge $d$ include, for each sender $a$, a time increment from edge $d-1$ at which sender $a$ would send edge $d$, and an $A-1$ length vector indicating which nodes would receive edge $d$ if it were sent by sender $a$.  The data generated for edge $d$ under the HEM corresponds to the sender that would send edge $d$ the soonest---at the smallest time increment from edge $d-1$. The receivers of edge $d$ generated under the HEM correspond to those receivers to which the sender with the minimum time increment would have sent edge $d$. We explain these steps in more detail below. For hyperedges that involve one receiver and one or more senders, we treat $a_d$ to be an indicator vector of senders $\boldsymbol{a}_d = (u_{dr})_{a=1}^{A}$ and $r_d$ to be the single receiver, and then follow the alternative generative process provided in Appendix A, which we derive as a simple reversal of the process used for multicasts (i.e., one sender and multiple receivers).

	\subsection{Candidate receivers}\label{subsec: Tie}
	For every possible sender--receiver pair $(a,r)_{a \neq r}$, we define the ``receiver intensity"---an approximate logit of the probability that edge $d$ is being sent from sender $a$ to receiver $r$---as a linear combination of statistics relevant to the receiver selection process:
	\begin{equation}
		\lambda_{adr} = {\boldsymbol{b}}^{\top}\boldsymbol{x}_{adr},
	\end{equation}
	where $\boldsymbol{b}$ is a $P$--dimensional vector of coefficients and $\boldsymbol{x}_{adr}$ is a set of receiver selection features. Such features can capture common network processes that include, but are not limited to, popularity, reciprocity, and transitivity, as well as the effects of attribute features defined on the sender, receivers, and/or sender-receiver pairs (e.g., the gender of the sender and/or receiver, an indicator of whether the sender is a supervisor of the receiver's). In addition, we include an intercept term to account for the average (or baseline) number of receivers. We place a Normal prior $\boldsymbol{b} \sim N(\boldsymbol{\mu}_b, \Sigma_b)$.

	A basic assumption of the HEM is that, at any given time, everyone is planning to send a document to one or more receivers. So for each edge $d$ and each sender $a$, the HEM draws a list of candidate receivers that would be receivers on edge $d$ if $a$ were the sender. For an edge $d$, we first define an $A\times A$ matrix $\boldsymbol{u}_d$ where the $a^{th}$ row denotes sender $a$'s receiver vector of zeros and 1's---i.e., 1's indicate the nodes to which $a$ intends to send edge $d$. We then assume that each receiver vector $\boldsymbol{u}_{ad}$ comes from a modification of the multivariate Bernoulli (MB) distribution \citep{dai2013multivariate}---a model that has been used to model graphs in which the state of each edge indicator is drawn independently from an edge-specific Bernoulli distribution. In order to avoid drawing edges with no receivers, we define a probability measure ``MB$_{G}$" motivated by the Gibbs measure \citep{fellows2017removing}. The probability measure we define amounts to a non-empty Gibbs measure, in which the all-zero vector is excluded from the support of the multivariate Bernoulli distribution. As a result, this measure helps us to 1) allow a sender to select multiple receivers for a single edge, 2) force the sender to select at least one receiver, and 3) ensure a tractable normalizing constant for the receiver selection distribution. To be specific, we draw a binary vector $\boldsymbol{u}_{ad}= (u_{ad1},
	\ldots, u_{adA})$ 
	\begin{equation} \boldsymbol{u}_{ad}  \sim
		\mbox{MB}_{G}(\boldsymbol{\lambda}_{ad}),
	\end{equation}
	where $\boldsymbol{\lambda}_{ad}= (\lambda_{ad1},
	\ldots, \lambda_{adA})$. In particular, we define $\mbox{MB}_{G}(\boldsymbol{\lambda}_{ad})$ as
	\begin{equation}
		\begin{aligned}
			&\Pr(\boldsymbol{u}_{ad}|\boldsymbol{b}, \boldsymbol{x}_{ad}) = \frac{1}{Z(\boldsymbol{\lambda}_{ad})}\exp\Big(\mbox{log}\big(\text{I}( \lVert \boldsymbol{u}_{ad}\rVert_1 > 0 )\big) + \sum_{r\neq a} \lambda_{adr}u_{adr}\Big) ,
		\end{aligned}
		\label{eqn:Gibbs}
	\end{equation}
	where $Z(\boldsymbol{\lambda}_{ad})= \prod_{r \neq a} \big(\mbox{exp}(\lambda_{adr}) + 1\big)-1$ is the normalizing constant and $\lVert \cdot \rVert_1$ is the $l_1$--norm. Again, this is approximately equivalent to assuming that each $u_{adr}$ is drawn with the probability of 1 being logit($\lambda_{adr}$), excluding the case when $u_{adr}=0$ for all $r \in [A]$. This allows efficient inference since we can derive a closed-form normalizing constant which does not require brute-force summation over the support of $\boldsymbol{u}_{ad}$ (i.e., $\forall \boldsymbol{u}_{ad} \in [0,1]^A$). We provide detailed derivation steps for the normalizing constant $Z(\boldsymbol{\lambda}_{ad})$ in Appendix B.

	\subsection{Candidate timestamps}\label{subsec:Time}
	Similarly, for each sender and edge combination, the HEM draws a candidate time at which the edge would be sent. The timing rate for sender $a$ and edge $d$ is
	\begin{equation}
		\mu_{ad} = g^{-1}(\boldsymbol{\eta}^\top \boldsymbol{y}_{ad}),
	\end{equation}
	where $\boldsymbol{\eta}$ is a $Q$--dimensional vector of coefficients with a Normal prior $\boldsymbol{\eta} \sim N(\boldsymbol{\mu}_\eta,\Sigma_\eta)$, $\boldsymbol{y}_{ad}$ is a set of event timing features---covariates that could affect timestamps of the edge, and $g(\cdot)$ is the appropriate link function such as identity, log, or inverse.

>>>>

		\begin{abstract}
			We introduce the hyperedge event model (HEM)---a generative model for events that can be represented as directed edges with one sender and one or more receivers or one receiver and one or more senders. We
			integrate a dynamic version of the exponential random graph model (ERGM) of edge structure with a survival model for event timing to jointly understand who interacts with whom, and when. The HEM offers three innovations with respect to the literature---first, it extends a growing class of dynamic network models to model hyperedges. The current state of the art approach to dealing with hyperedges is to inappropriately break them into separate edges/events. Second, the HEM involves a novel receiver-selection distribution that is based on established edge formation models, but assures non-empty receiver lists. Third, HEM integrates separate, but interacting, equations governing edge formation and event timing. We use the HEM to analyze emails sent among department managers in Montgomery County government in North Carolina. Our application demonstrates that the model is effective at predicting and explaining time-stamped network data involving edges with multiple receivers. We present an out-of-sample prediction experiment to illustrate how researchers can select between different specifications of the model. 
		\end{abstract}

@@

	In this section we define the HEM by describing a process according to which hyperedge events are generated in continuous time. Data generated under the HEM consists of $D$ unique edges. A single edge indexed by $d \in [D]$, where $[D]$ denotes a categorical set with $D$ levels $[D] = \{1,\ldots,D\}$, is represented by the three components: the sender $a_d \in [A]$, an indicator vector of receivers $\boldsymbol{r}_d = (u_{dr})_{r=1}^{A}$, such that $u_{dr}=1$ if $r$ is a receiver of edge $d$ and 0 otherwise, and the timestamp $t_d \in (0, \infty)$. For simplicity, we assume that edges are ordered by time such that $t_d \leq t_{d+1}$. While the model can be applied for two types of hyperedges---edges with (1) one sender and one or more receivers, and (2) one or more senders and one receiver---here we only present the generative process for those involving one sender and one or more receivers (i.e., multicast). One notable feature of the HEM generative process is that we draw auxiliary variables that serve as candidate data. Data is generated from the HEM through a sampling process applied to the auxiliary variables. The auxiliary variables drawn for edge $d$ include, for each sender $a$, a time increment from edge $d-1$ at which sender $a$ would send edge $d$, and an $A-1$ length vector indicating which nodes would receive edge $d$ if it were sent by sender $a$.  The data generated for edge $d$ under the HEM corresponds to the sender that would send edge $d$ the soonest---at the smallest time increment from edge $d-1$. The receivers of edge $d$ generated under the HEM correspond to those receivers to which the sender with the minimum time increment would have sent edge $d$. We explain these steps in more detail below. For hyperedges that involve one receiver and one or more senders, we treat $a_d$ to be an indicator vector of senders $\boldsymbol{a}_d = (u_{dr})_{a=1}^{A}$ and $r_d$ to be the single receiver, and then follow the alternative generative process provided in Appendix A, which we derive as a simple reversal of the process used for multicasts (i.e., one sender and multiple receivers).

	\subsection{Candidate receivers}\label{subsec: Tie}
	For every possible sender--receiver pair $(a,r)_{a \neq r}$, we define the ``receiver intensity"---the conditional logit of the probability that edge $d$ is sent from sender $a$ to receiver $r$---as a linear combination of statistics relevant to the receiver selection process:
	\begin{equation}
		\lambda_{adr} = {\boldsymbol{b}}^{\top}\boldsymbol{x}_{adr},
	\end{equation}
	where $\boldsymbol{b}$ is a $P$--dimensional vector of coefficients and $\boldsymbol{x}_{adr}$ is a set of receiver selection features. $\lambda_{adr}$ is the logit of the probability that edge $d$ is sent from sender $a$ to receiver $r$ conditional upon there being at least one other receiver to which sender $a$ sends edge $d$. In the case that there are no other receivers to which sender $a$ sends edge $d$, the logit of the probability that sender $a$ sends edge $d$ to receiver $r$ is $\infty$.  The features in Equation 2.1 can capture common network processes that include, but are not limited to, popularity, reciprocity, and transitivity, as well as the effects of attribute features defined on the sender, receivers, and/or sender-receiver pairs (e.g., the gender of the sender and/or receiver, an indicator of whether the sender is a supervisor of the receiver's). In addition, we include an intercept term to account for the average (or baseline) number of receivers. We place a Normal prior $\boldsymbol{b} \sim N(\boldsymbol{\mu}_b, \Sigma_b)$.

	As we describe in greater detail below, under the HEM the sender of each edge is selected to be the sender that would send the respective edge with the greatest urgency. As such, under the HEM we assume that each sender would plan on sending each edge to some set of recipients at some time point. For each edge $d$ and each sender $a$, the HEM draws a list of candidate receivers that would be receivers on edge $d$ if $a$ were the sender. For an edge $d$, we first define an $A\times A$ matrix $\boldsymbol{u}_d$ where the $a^{th}$ row denotes sender $a$'s receiver vector of zeros and 1's---i.e., 1's indicate the nodes to which $a$ intends to send edge $d$. We then assume that each receiver vector $\boldsymbol{u}_{ad}$ comes from a modification of the multivariate Bernoulli (MB) distribution \citep{dai2013multivariate}---a model that has been used to model graphs in which the state of each edge indicator is drawn independently from an edge-specific Bernoulli distribution. In order to avoid drawing edges with no receivers, we define a probability measure ``MB$_{G}$" motivated by the Gibbs measure \citep{fellows2017removing}. The probability measure we define amounts to a non-empty Gibbs measure, in which the all-zero vector is excluded from the support of the multivariate Bernoulli distribution. As a result, this measure helps us to 1) allow a sender to select multiple receivers for a single edge, 2) force the sender to select at least one receiver, and 3) ensure a tractable normalizing constant for the receiver selection distribution. To be specific, we draw a binary vector $\boldsymbol{u}_{ad}= (u_{ad1},
	\ldots, u_{adA})$ 
	\begin{equation} \boldsymbol{u}_{ad}  \sim
		\mbox{MB}_{G}(\boldsymbol{\lambda}_{ad}),
	\end{equation}
	where $\boldsymbol{\lambda}_{ad}= (\lambda_{ad1},
	\ldots, \lambda_{adA})$. In particular, we define $\mbox{MB}_{G}(\boldsymbol{\lambda}_{ad})$ as
	\begin{equation}
		\begin{aligned}
			&\Pr(\boldsymbol{u}_{ad}|\boldsymbol{b}, \boldsymbol{x}_{ad}) = \frac{1}{Z(\boldsymbol{\lambda}_{ad})}\exp\Big(\mbox{log}\big(\text{I}( \lVert \boldsymbol{u}_{ad}\rVert_1 > 0 )\big) + \sum_{r\neq a} \lambda_{adr}u_{adr}\Big) ,
		\end{aligned}
		\label{eqn:Gibbs}
	\end{equation}
	where $Z(\boldsymbol{\lambda}_{ad})= \prod_{r \neq a} \big(\mbox{exp}(\lambda_{adr}) + 1\big)-1$ is the normalizing constant and $\lVert \cdot \rVert_1$ is the $l_1$--norm. Again, this is approximately equivalent to assuming that each $u_{adr}$ is drawn with the probability of 1 being logit($\lambda_{adr}$), excluding the case when $u_{adr}=0$ for all $r \in [A]$. This allows efficient inference since we can derive a closed-form normalizing constant which does not require brute-force summation over the support of $\boldsymbol{u}_{ad}$ (i.e., $\forall \boldsymbol{u}_{ad} \in [0,1]^A$). We provide detailed derivation steps for the normalizing constant $Z(\boldsymbol{\lambda}_{ad})$ in Appendix B.

	\subsection{Candidate timestamps}\label{subsec:Time}
	To generate timing for each edge, the HEM first draws a candidate time at which the edge would be sent given the candidate sender and receiver combinations. The timing rate for sender $a$ and edge $d$ is
	\begin{equation}
		\mu_{ad} = g^{-1}(\boldsymbol{\eta}^\top \boldsymbol{y}_{ad}),
	\end{equation}
	where $\boldsymbol{\eta}$ is a $Q$--dimensional vector of coefficients with a Normal prior $\boldsymbol{\eta} \sim N(\boldsymbol{\mu}_\eta,\Sigma_\eta)$, $\boldsymbol{y}_{ad}$ is a set of event timing features---covariates that could affect timestamps of the edge, and $g(\cdot)$ is the appropriate link function such as identity, log, or inverse.
